Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-05 talk/dls2014/paper/paper.tex
start equalising the style of writing

\section{High-level overview}

@@

A naive approach to providing complete isolation between threads is to
partition the virtual memory of a process into $N$ segments, one per
thread.  Each segment then holds a complete copy of the memory logically
seen by the program.  All threads read and write data inside their own
segment; exchanges between segments occur only at commit-time.  So far,
this is a model that would also work for distributed transactional
memory.

Obviously, if we are in a shared-memory setting, this model is extremely
wasteful: at a given time, most objects are identical in all, or most,
segments.  They differ only in segments that have modified but not yet
committed the object.  So we use a trick of the Memory Management Unit
(MMU) of the CPU.  At the coarse unit of one page of memory (4096
bytes), we use remappings so that, in general, pages within segments are
\emph{shared} with each other if they would contain a fully identical
copy.  Conversely, when one segment needs to write to one object
contained in a page, the complete page is first \emph{privatised:} it is
duplicated and the copy is remapped so that it will be visible in this
segment (only).  This process is illustrated in
Figure~\ref{fig:Page-Remapping}.

The result that we get is that the virtual address space of the process
is bloated -- by a factor $N$ -- but the real memory usage is not --
because most pages are shared with their copies in most other segments.

@@

What we described so far has no mesurable performance impact by
itself, and it is missing conflict detection to be considered a
full-fledged STM system.  This requires adding \emph{barriers} to the
program to register the objects into the read or write set before
reading or writing to them\footnote{Barriers added by an automatic and
local program transformation~\cite{felber07}}.  Furthermore, we need
to describe the commit protocol.

@@

\item [{Read~Barrier:}] A major contribution to our performance results
  is the simplicity of the read barrier: a large program (like an
  interpreter for a complex language) contains a number of reads that
  typically exceeds many times the number of writes.  We managed to
  design the read barrier so that it is done by one memory write
  (setting a flag) into non-shared memory.  This is done with a
  segment-local array of flags indexed by a number derived from the $SO$
  of the object.  Unlike other STM systems, the read barrier does not
  have to find the private version of some object -- our two-step
  address translation automatically resolves the reference to the
  private version on every access anyway.  This means the read barrier
  is not constrained to execute before the actual read -- both the
  compiler and the CPU are free to reorder them.

>>>>

\section{High-Level Overview}

@@

Our STM system fundamentally builds on threads. Transactions run in
their context, and there can be only one transaction per thread at a
time. Transactions run in parallel because these threads can run in
parallel. In order to isolate transactions from each other, we look
at isolating the memory accesses from threads.

A naive approach to providing this complete isolation between threads
is to partition the virtual memory of a process into $N$ segments, one
for each thread currently executing.  Each segment then holds a
complete copy of the memory logically seen by the program.  All
threads read and write data inside their own segment; exchanges
between segments occur only when transactions commit.  So far, this is a
model that would also work for distributed transactional memory.

Clearly, if we are in a shared-memory setting, this model is extremely
wasteful: at any given time, most objects are identical in all, or
most, segments.  They differ only in segments that have modified but
not yet committed these objects.  So we use a trick of the Memory
Management Unit (MMU) of the CPU.  The MMU maps pages of virtual to
physical memory.  By changing this page mapping, at the coarse unit of
one page of memory (4096 bytes), we can share memory that would
otherwise exist as identical copies in all segments.

In the common case, where objects are identical in all segments, we
can thus make these pages \emph{shared}. Conversely, when one thread~/
segment wants to modify an object, all pages of this object need to be
\emph{privatised} for this segment. This involves duplicating the
pages and changing the page mapping so that the private pages are only
accessible by this segment and full isolation is guaranteed.  This
process is illustrated in Figure~\ref{fig:Page-Remapping}.

The result is that the virtual address space of the process is bloated
by a factor $N$ -- but the real memory usage is not, because most
pages are shared with their identical copies in other segments.

@@

What we described so far has no mesurable performance impact by
itself, and it is missing conflict detection to be considered a
full-fledged STM system.  This requires adding \emph{barriers} to the
program to register the objects in the read or write set before
reading or writing to them\footnote{Barriers added by an automatic and
local program transformation~\cite{felber07}}.  Furthermore, we need
to describe the commit protocol.

@@

\item [{Read~Barrier:}] A major contribution to our performance results
  is the simplicity of the read barrier: a large program (like an
  interpreter for a complex language) contains a number of reads that
  typically exceeds many times the number of writes.  The described
  approach allows the read barrier to be a single, unconditional memory
  write into segment-local memory.
 We managed to
  design the read barrier so that it is done by one memory write
  (setting a flag) into non-shared memory.  This is done with a
  segment-local array of flags indexed by a number derived from the $SO$
  of the object.  Unlike other STM systems, the read barrier does not
  have to find the private version of some object -- our two-step
  address translation automatically resolves the reference to the
  private version on every access anyway.  This means the read barrier
  is not constrained to execute before the actual read -- both the
  compiler and the CPU are free to reorder them.
